Efficient FPGA Implementation of\\ Digit Parallel Online Arithmetic Operators

### 2014-10-15 FPT2014/bare_conf.tex
change file format for figures

% Graph package
\ifx\pdfoutput\undefined
\usepackage{graphicx}
\else
\usepackage[pdftex]{graphicx}
\fi
\usepackage{epstopdf}

@@

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
\usepackage[caption=false,font=footnotesize]{subfig}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle

@@

%In summary, the main contributions of this paper are efficient methods to map online adders and multipliers to modern FPGAs. The performance improvements in terms of silicon area and operating frequency with these new architectures mean that using online operators is much more attractive for an end user. The benefits of these operators .

@@

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.\vspace{-0.5ex}
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible only if a redundant number system is used. Normally there are two most commonly used redundant number representations: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. With SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the binary number $0.011$ can be represented in SD form as $0.1\overline{1}1$, $0.10\overline{1}$ or $0.011$ among many other possible representations.\vspace{-0.5ex}
%
\vspace{-.5ex}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/SDAdder2.pdf}
    \vspace{-7ex}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
    \vspace{-2ex}
  \label{Fig:Radix2SD_adder}
\end{figure}

@@

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the two FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea of building the online adder based on the logic block (\texttt{LB}) as highlighted within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the two 3:2 compressors can be greatly reduced.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan2.pdf}
	\vspace{-4ex}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The gray background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\vspace{-4ex}
	\label{Fig:PreviousWork}
\end{figure}

@@

\begin{figure*}[tbp]
	\centering
	\includegraphics[width=.85\textwidth]{./Figures/SDadder_FPGA_New_4digit2.pdf}
	\vspace{-3ex}
	\caption{Alternative structure of online adder. Left: an example of 4-digit online adder. The shaded part refers to the two logic blocks (LBs) that can be mapped onto one slice. Right: one LB. The dotted boxes outline the logic that can be mapped onto each LUT and the corresponding fast carry logic.}
	\vspace{-2ex}	
	\label{Fig:ModifiedOA}
\end{figure*}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.49\textwidth]{./Figures/SDadder_FastCarry_New2.pdf}
	\vspace{-5ex}
	\caption{Implementation of two logic blocks (LBs) in one FPGA slice which contains four 6-LUTs. NC stands for ``Not Care''.}
	\vspace{-2ex}	
	\label{Fig:SliceNew}
\end{figure}

@@

\subsection{FPGA Implementation}\vspace{-0.5ex}
We now present an area efficient implementation of Algorithm~\ref{Algorithm:OM_DigitParallel} on FPGAs. The general structure of a 4-digit \texttt{OM} using the proposed algorithm is shown in Fig.~\ref{Fig:PMStructure}(a), and the structure of a single stage $j$ is shown in Fig.~\ref{Fig:PMStructure}(b) with the word-lengths of all signals labelled. Area optimizations can be performed on all modules. It can be seen that in each stage, an $(N+2)$ digit online adder is used to derive $W_{[j]}$. Our proposed online adder architecture can be employed. The selection logic takes 3 input digits (6 bits) and generates 1 output digit (2 bits). Hence it can be implemented using two 6-LUTs for each output bit. Similarly the generation of $P_{[j]}$ can also be implemented using one 6-LUT, since only the integer digits of $W_{[j]}$ need to be modified due to the selection of $Z_{[j]}$. In addition, the structure of Stage 1 in Fig~\ref{Fig:PMStructure}(a) can be further optimized by removing the online adder, because $P_{[0]}=0$.\vspace{-0.5ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_Structure.pdf}
	\vspace{-3ex}
	\caption{(a) Structure of a 4-digit online multiplier using the proposed algorithm. (b) Structure of one stage. The word-length of all signals are labelled in terms of the number of digits. $N$ denotes the word-length of the input signals.}
	\label{Fig:PMStructure}
	\vspace{-2ex}
\end{figure}

The blocks that generate partial products $Xy_j$ can be incorporated into the online adder for further area reduction. In the online adder structure as presented in Fig.~\ref{Fig:SliceNew}, we notice that only four inputs per LUT are used. This leaves at least two available inputs per LUT. Originally in order to generate one digit partial product, one digit of inputs $X$ and $Y$ are required respectively, as indicated in Fig.~\ref{Fig:PartialProduct}. Alternatively, instead of using extra logic to generate $Xy_j$, it can be merged into the corresponding LUTs in the online adder by fully utilizing all six LUT inputs, as shown on the right of Fig.\ref{Fig:PartialProduct}. Notice that this approach is only available for FPGAs with 6-LUTs.\vspace{-0.5ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.4\textwidth]{./Figures/PartialProduct.pdf}
	\vspace{-2.5ex}
	\caption{Left: Direct implementation with extra logic to generate partial products. Right: Combining the logic blocks that generate partial products into the Online Adder by fully utilizing all the inputs of the 6-LUTs. In this case, further area reduction can be achieved.}
	\label{Fig:PartialProduct}
	\vspace{-1.5ex}
\end{figure}

@@

In comparison, the \texttt{OM} offers the flexibility to simplify the structure corresponding to the required precision. This is possible because in an \texttt{OM}, the product digits are generated initially from the MSD, and there is no carry propagation from the LSD to the MSD with the employment of the redundant number system. This will potentially lead to a more area efficient design. For example, the modified structure diagram of a 4-digit \texttt{OM} is illustrated in Fig.~\ref{Fig:PMStructure_MSDhalf}(a). The word-length of signals within each stage can be correspondingly reduced, as shown in Fig.~\ref{Fig:PMStructure_MSDhalf}(b). It can be seen that instead of keeping identical word-length throughout the stages (in Fig.~\ref{Fig:PMStructure}), in the modified structure the signal word-lengths depend on the stage number $j$ and fewer bits of signals are needed for LSD stages.\vspace{-3ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_MSDhalf.pdf}
  \vspace{-2ex}
	\caption{(a) Modified structure of a 4-digit online multiplier which only generates the most significant 4-digit result. (b) Structure diagram of stage $j$, with the word-lengths of all internal signals labelled.}
	\label{Fig:PMStructure_MSDhalf}
	\vspace{-2ex}
\end{figure}

@@

%\section{Results}

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

@@

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

>>>>

% Graph package
%\ifx\pdfoutput\undefined
\usepackage{graphicx}
%\else
%\usepackage[pdftex]{graphicx}
%\fi
%\usepackage{epstopdf}

@@

\usepackage[caption=false,font=footnotesize]{subfig}

@@

f

@@

@@

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.\vspace{-0.5ex}
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.eps}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible only if a redundant number system is used. Normally there are two most commonly used redundant number representations: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. With SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the binary number $0.011$ can be represented in SD form as $0.1\overline{1}1$, $0.10\overline{1}$ or $0.011$ among many other possible representations.\vspace{-0.5ex}
%
\vspace{-.5ex}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/SDAdder.eps}
    \vspace{-7ex}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
    \vspace{-2ex}
  \label{Fig:Radix2SD_adder}
\end{figure}

@@

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the two FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea of building the online adder based on the logic block (\texttt{LB}) as highlighted within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the two 3:2 compressors can be greatly reduced.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan.eps}
	\vspace{-4ex}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The gray background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\vspace{-4ex}
	\label{Fig:PreviousWork}
\end{figure}

@@

\begin{figure*}[tbp]
	\centering
	\includegraphics[width=.85\textwidth]{./Figures/SDadder_FPGA_New_4digit.eps}
	\vspace{-3ex}
	\caption{Alternative structure of online adder. Left: an example of 4-digit online adder. The shaded part refers to the two logic blocks (LBs) that can be mapped onto one slice. Right: one LB. The dotted boxes outline the logic that can be mapped onto each LUT and the corresponding fast carry logic.}
	\vspace{-2ex}	
	\label{Fig:ModifiedOA}
\end{figure*}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.49\textwidth]{./Figures/SDadder_FastCarry_New.eps}
	\vspace{-5ex}
	\caption{Implementation of two logic blocks (LBs) in one FPGA slice which contains four 6-LUTs. NC stands for ``Not Care''.}
	\vspace{-2ex}	
	\label{Fig:SliceNew}
\end{figure}

@@

\subsection{FPGA Implementation}\vspace{-0.5ex}
We now present an area efficient implementation of Algorithm~\ref{Algorithm:OM_DigitParallel} on FPGAs. The general structure of a 4-digit \texttt{OM} using the proposed algorithm is shown in Fig.~\ref{Fig:PMStructure}(a), and the structure of a single stage $j$ is shown in Fig.~\ref{Fig:PMStructure}(b) with the word-lengths of all signals labelled. Area optimizations can be performed on all modules. It can be seen that in each stage, an $(N+2)$ digit online adder is used to derive $W_{[j]}$. Our proposed online adder architecture can be employed. The selection logic takes 3 input digits (6 bits) and generates 1 output digit (2 bits). Hence it can be implemented using two 6-LUTs for each output bit. Similarly the generation of $P_{[j]}$ can also be implemented using one 6-LUT, since only the integer digits of $W_{[j]}$ need to be modified due to the selection of $Z_{[j]}$. In addition, the structure of Stage 1 in Fig~\ref{Fig:PMStructure}(a) can be further optimized by removing the online adder, because $P_{[0]}=0$.\vspace{-0.5ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_Structure.eps}
	\vspace{-3ex}
	\caption{(a) Structure of a 4-digit online multiplier using the proposed algorithm. (b) Structure of one stage. The word-length of all signals are labelled in terms of the number of digits. $N$ denotes the word-length of the input signals.}
	\label{Fig:PMStructure}
	\vspace{-2ex}
\end{figure}

The blocks that generate partial products $Xy_j$ can be incorporated into the online adder for further area reduction. In the online adder structure as presented in Fig.~\ref{Fig:SliceNew}, we notice that only four inputs per LUT are used. This leaves at least two available inputs per LUT. Originally in order to generate one digit partial product, one digit of inputs $X$ and $Y$ are required respectively, as indicated in Fig.~\ref{Fig:PartialProduct}. Alternatively, instead of using extra logic to generate $Xy_j$, it can be merged into the corresponding LUTs in the online adder by fully utilizing all six LUT inputs, as shown on the right of Fig.\ref{Fig:PartialProduct}. Notice that this approach is only available for FPGAs with 6-LUTs.\vspace{-0.5ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.4\textwidth]{./Figures/PartialProduct.eps}
	\vspace{-2.5ex}
	\caption{Left: Direct implementation with extra logic to generate partial products. Right: Combining the logic blocks that generate partial products into the Online Adder by fully utilizing all the inputs of the 6-LUTs. In this case, further area reduction can be achieved.}
	\label{Fig:PartialProduct}
	\vspace{-1.5ex}
\end{figure}

@@

In comparison, the \texttt{OM} offers the flexibility to simplify the structure corresponding to the required precision. This is possible because in an \texttt{OM}, the product digits are generated initially from the MSD, and there is no carry propagation from the LSD to the MSD with the employment of the redundant number system. This will potentially lead to a more area efficient design. For example, the modified structure diagram of a 4-digit \texttt{OM} is illustrated in Fig.~\ref{Fig:PMStructure_MSDhalf}(a). The word-length of signals within each stage can be correspondingly reduced, as shown in Fig.~\ref{Fig:PMStructure_MSDhalf}(b). It can be seen that instead of keeping identical word-length throughout the stages (in Fig.~\ref{Fig:PMStructure}), in the modified structure the signal word-lengths depend on the stage number $j$ and fewer bits of signals are needed for LSD stages.\vspace{-3ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/ParallelMult_MSDhalf.eps}
  \vspace{-2ex}
	\caption{(a) Modified structure of a 4-digit online multiplier which only generates the most significant 4-digit result. (b) Structure diagram of stage $j$, with the word-lengths of all internal signals labelled.}
	\label{Fig:PMStructure_MSDhalf}
	\vspace{-2ex}
\end{figure}

@@

@@
